The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2013-01-14 script/en/chapters/numbertheory.tex
Several minor changes (for consistency and correctness), mostly in chap 4.

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\section[Divisibility, modulus and remainder classes]
           {Divisibility, modulus and remainder classes\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can have a
    playful view at the calculation with congruences, discussed in this and
    the next chapter (see NT learning unit 2.1, pages 2-9/40).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix~\ref{s:appendix-Learn-NT}.\\
    CT2\index{CrypTool 2} contains a visualisation of these methods within
    the tutorial ``{\bf World of Primes}''.
}
\index{Modulus} \index{Divisibility}

@@

\newpage
\begin{example}{:}
\begin{itemize}
\item[] Remainder class $0$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n; \; n \in \mathbb{Z} \} = \{ \dots, -16, -12, -8, -4, 0, 4, 8, 12, 16, \dots \}$
\item[] Remainder class $3$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n + 3;\; n \in \mathbb{Z} \} = \{ \dots, -13, -9, -5, -1, 3, 7, 11, 15, \dots \}$
\end{itemize}
\end{example}
As only the remainders $0, 1, 2, \cdots, m-1$ are possible for division modulo $m$, modular arithmetic works with finite sets. 
For each modulo $m$ there are precisely $m$ remainder classes.

@@

\begin{example}{:}
\begin{itemize}
   \item[] $19 \equiv 12 {\rm ~(mod~} 7)$,         
           because the remainders are equal:  $19 / 7 = 2$ remainder $5$  and  $12 / 7 = 1$ remainder $5$.
   \item[] $23103 \equiv 0 {\rm ~(mod~} 453)$, because $23103 / 453 = 51$ remainder $0$  and  $0 / 453 = 0$ remainder $0$.
\end{itemize}
\end{example}

\begin{theorem}\label{thm-zth-div}
$a \equiv b$ (mod $m$) if and only if, the difference $(a - b)$ is divisible by $m$, i.e. if $q\in 
\mathbb{Z}$ exists with $ (a-b)=q*m.$
\end{theorem}
These two statements are therefore equivalent.

@@

\begin{example}{ Example of equivalent statements:}\\
$35 \equiv 11$ (mod $3) \Longleftrightarrow  35 - 11 \equiv 0$ (mod $3)$, 
where $35 - 11 = 24$ is divisible by $3$ without remainder while $35:3$ and $11:3$
leave the remainder $2$.
\end{example}

\begin{remark}{:}\\
The above equivalence does not apply to the sum $(a + b)$!
\end{remark}

\begin{example}{:}\\
$11 \equiv 2$ (mod $3$), therefore $11 - 2 = 9 \equiv 0$ (mod $3$);
but $11 + 2 = 13$ is not divisible by $3$.
The statement in theorem~\ref{thm-zth-div} does not even apply to sums in one direction.
It is correct for sums only if the remainder is 0 and only in the following direction:
If a divisor divides both summands with no remainder, it also divides the sum with no remainder.
\end{example}

@@

% --------------------------------------------------------------------------
\vskip +10pt
\hypertarget{hohpot}{}
\subsection{Fast calculation of high powers} \label{hohpot}\index{Power}
RSA encryption and decryption\footnote{See chapter~\ref{rsabeweis} (Proof of the
RSA procedure with Euler-Fermat) \index{RSA} and chapter~\ref{rsaconcrete} (The
RSA procedure with actual numbers).} entails calculating high powers modulo
$m$.  For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit
long integer\index{Long integer} number range provided we calculate $a^n$ by
actually multiplying a with itself $n$ times in line with the definition. In
the case of extremely large numbers, even a fast computer chip would take
longer than the age of the universe to calculate a single exponential. Luckily,
there is an extremely effective shortcut for calculating exponentials (but not
for calculating logarithms).

@@

% --------------------------------------------------------------------------
\subsection{How many private RSA keys $d$ are there modulo 26}

@@

\begin{table}[ht]
{\textmd \small
\begin{center}  
\begin{tabular}{|p{16 pt}||@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|c|}
\hline
$a \setminus i$   & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 & 20 & 21 & 22 & 23 & ord \\
\hline
\hline
1    & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1 & 1    \\
\hline
2 & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2 & --    \\
\hline
3 & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3 & 11   \\
\hline
4  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & --  \\
\hline
5 & 5 & 25 & 33 & 27 & 43 & 31 & 17 & 39 & 11  & 9 & 45 & 41 & 21 & 13 & 19  & 3 & 15 & 29  & 7 & 35 & 37  & 1  & 5 & 22  \\
\hline
6 & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & -- \\
\hline
7 & 7  & 3 & 21  & 9 & 17 & 27  & 5 & 35 & 15 & 13 & 45 & 39 & 43 & 25 & 37 & 29 & 19 & 41 & 11 & 31 & 33  & \textbf{1}  & 7 & 22 \\
\hline
8 & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & --  \\
\hline
9 & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 11  \\
\hline
10 & 10  & 8 & 34 & 18 & 42  & 6 & 14  & 2 & 20 & 16 & 22 & 36 & 38 & 12 & 28  & 4 & 40 & 32 & 44 & 26 & 30 & 24 & 10 & --  \\
\hline 
11 & 11 & 29 & 43 & 13  & 5  & 9  & 7 & 31 & 19 & 25 & 45 & 35 & 17  & 3 & 33 & 41 & 37 & 39 & 15 & 27 & 21  & 1 & 11 & 22 \\
\hline
12 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12 & -- \\
\hline
13 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 11  \\
\hline
14 & 14 & 12 & 30  & 6 & 38 & 26 & 42 & 36 & 44 & 18 & 22 & 32 & 34 & 16 & 40  & 8 & 20  & 4 & 10  & 2 & 28 & 24 & 14 & -- \\
\hline
15 & 15 & 41 & 17 & 25  & 7 & 13 & 11 & 27 & 37  & 3 & 45 & 31  & 5 & 29 & 21 & 39 & 33 & 35 & 19  & 9 & 43  & 1 & 15 & 22 \\
\hline
16 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & -- \\
\hline
17 & 17 & 13 & 37 & 31 & 21 & 35 & 43 & 41  & 7 & 27 & 45 & 29 & 33  & 9 & 15 & 25 & 11  & 3  & 5 & 39 & 19  & 1 & 17 & 22 \\
\hline
18 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18 & -- \\
\hline
19  & 19 & 39  & 5  & 3 & 11 & 25 & 15  & 9 & 33 & 29 & 45 & 27  & 7 & 41 & 43 & 35 & 21 & 31 & 37 & 13 & 17  & 1 & 19 & 22 \\
\hline
20  & 20 & 32 & 42 & 12 & 10 & 16 & 44  & 6 & 28  & 8 & 22 & 26 & 14  & 4 & 34 & 36 & 30  & 2 & 40 & 18 & 38 & 24 & 20 & --  \\
\hline
21 & 21 & 27 & 15 & 39 & 37 & 41 & 33  & 3 & 17 & 35 & 45 & 25 & 19 & 31  & 7  & 9  & 5 & 13 & 43 & 29 & 11  & 1 & 21 & 22  \\
\hline
22 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & -- \\
\hline
23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & -- \\
\hline 
\end{tabular}
\end{center}
}
\caption{Values of $a^i{\rm ~mod~}46, 1\leq a,i<23$}
\label{SrcArith3c}
\label{expmod46}
\end{table}

\vskip +10 pt
\begin{theorem}\label{thm-zth-ordp}
  Given a modulus $n$ and a number $a$, relative prime to $n$ the
  following holds: \\
  The set $\{ a^i\; (\mbox{mod }n) |\; i = 1,\dots,J(n)\}$ equals the
  multiplicative group $Z_n^*$ if and only if $ord_n(a) = J(n)$.\footnote{%
    For prime moduli $p$ all $a$ with $0 < a < p$ are of order $J(p) = p - 1$.
    Compare table~\ref{expmod45} for an example. In this case $a^i (\mbox{mod }n)$ goes
    through all the values $1,\dots,p-1$. Exhausting all possible values of
    the set is an important cryptographic proposition (compare theorem
    \ref{thm-zth-exhperm}). This determines a permutation\index{Permutation}
    $\pi(p-1)$.}$^,$\footnote{%
    Table~\ref{expmod46} demonstrates that for composite moduli $n$ not all $a$ are of
    maximal order $J(n)$. In this example only $5,7,11,15,17,19\mbox{ and }21$
    are of order 22.
  }
\end{theorem}
\noindent The multiplicative group $Z_n^*$ contains all values from $1$ to $n-1$.

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
%\hypertarget{Chapter_ElementaryNT_10}{}
\hypertarget{RSABeweis}{}
\section{Proof of the RSA procedure with Euler-Fermat}
\index{RSA}\label{rsabeweis}
Using the Euler-Fermat theorem, we can ``prove'' the
RSA\footnote{The RSA procedure\index{RSA!RSA procedure} is the most common
asymmetric cryptography procedure. Developed in 1978 by Ronald Rivest,
Adi Shamir and Leonard Adleman, it can be used both for signatures and for
encryption.
Cryptographers always associate this procedure with the abbreviation
``\textbf{RSA}'' $-$ the following remark is meant with humor to show that
each letter combination can be used with several meanings: In Britain
the ``Royal Society for the encouragement of Arts, Manufactures \& Commerce''
is commonly known as the ``RSA''.
}
procedure in the group $\mathbb{Z}_n^*$.

>>>>

% ++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\newpage
\section[Divisibility, modulus and remainder classes]
           {Divisibility, modulus and remainder classes\footnotemark}
\footnotetext{%
    \index{NT, Learning Tool for Number Theory}%
    \index{Educational tool NT}%
    With the educational tool for number theory {\bf NT} you can have a
    playful view at the calculation with congruences, discussed in this and
    the next chapter (see NT learning unit 2.1, pages 2-9/40).\\
    NT can be called in CrypTool\index{CrypTool} via the menu path
    {\bf Indiv. Procedures \textbackslash{} Number Theory Interactive
    \textbackslash{} Learning Tool for Number Theory}.
    See appendix~\ref{s:appendix-Learn-NT}.\\
    CT2\index{CrypTool 2} contains a visualisation of these methods within
    the tutorial ``{\bf World of Primes}''.
}
\index{Modulus} \index{Divisibility}

@@

\newpage
\begin{example}{:}
\begin{compactitem}
\item[] Remainder class $0$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n; \; n \in \mathbb{Z} \} = \{ \dots, -16, -12, -8, -4, 0, 4, 8, 12, 16, \dots \}$
\item[] Remainder class $3$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n + 3;\; n \in \mathbb{Z} \} = \{ \dots, -13, -9, -5, -1, 3, 7, 11, 15, \dots \}$
\end{compactitem}
\end{example}
As only the remainders $0, 1, 2, \cdots, m-1$ are possible for division modulo $m$, modular arithmetic works with finite sets. 
For each modulo $m$ there are precisely $m$ remainder classes.

@@

\begin{example}{:}
\begin{compactitem}
   \item[] $19 \equiv 12 {\rm ~(mod~} 7)$,         
           because the remainders are equal:  $19 / 7 = 2$ remainder $5$  and  $12 / 7 = 1$ remainder $5$.
   \item[] $23103 \equiv 0 {\rm ~(mod~} 453)$, because $23103 / 453 = 51$ remainder $0$  and  $0 / 453 = 0$ remainder $0$.
\end{compactitem}
\end{example}

\begin{theorem}\label{thm-zth-div}
$a \equiv b$ (mod $m$) if and only if, the difference $(a - b)$ is divisible by $m$, i.e. if $q\in 
\mathbb{Z}$ exists with $ (a-b)=q*m.$
\end{theorem}
These two statements are therefore equivalent.\footnote{%
The above equivalence does apply only to the difference $(a - b)$, not to the
sum $(a + b)$!

\begin{example}{:}\\
$11 \equiv 2$ (mod $3$), therefore $11 - 2 = 9 \equiv 0$ (mod $3$);
but $11 + 2 = 13$ is not divisible by $3$.
\end{example}

\noindent The statement in theorem~\ref{thm-zth-div} does not even apply to
sums in one direction.
It is correct for sums only if the remainder is 0 and only in the following direction:
If a divisor divides both summands with no remainder, it also divides the sum with no remainder.
}

@@

\begin{example}{ of equivalent statements:}\\
$35 \equiv 11$ (mod $3) \Longleftrightarrow  35 - 11 \equiv 0$ (mod $3)$, 
where $35 - 11 = 24$ is divisible by $3$ without remainder while $35:3$ and $11:3$
leave the remainder $2$.
\end{example}

@@

% --------------------------------------------------------------------------
\vskip +10pt
\hypertarget{hohpot}{}
\subsection{Fast calculation of high powers} \label{hohpot}\index{Power}
RSA encryption and decryption\footnote{See chapter~\ref{rsabeweis} (``\nameref{rsabeweis}'') \index{RSA} and chapter~\ref{rsaconcrete} (``\nameref{rsaconcrete}'').} 
entails calculating high powers modulo $m$.
For example, the calculation ($100^5) {\rm ~(mod~} 3)$ exceeds the 32-bit
long integer\index{Long integer} number range provided we calculate $a^n$ by
actually multiplying a with itself $n$ times in line with the definition. In
the case of extremely large numbers, even a fast computer chip would take
longer than the age of the universe to calculate a single exponential. Luckily,
there is an extremely effective shortcut for calculating exponentials (but not
for calculating logarithms).

@@

% --------------------------------------------------------------------------
\subsection{How many private RSA keys \texorpdfstring{$d$}{d} are there modulo 26}
\label{L_nt_Num-of-d-mod-26}

@@

\begin{table}[ht]
{\textmd \small
\begin{center}  
\begin{tabular}{|p{16 pt}||@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|@{\:}r@{\:}|c|}
\hline
$a \setminus i$   & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & 14 & 15 & 16 & 17 & 18 & 19 & 20 & 21 & 22 & 23 & ord \\
\hline
\hline
1    & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1  & 1 & 1    \\
\hline
2 & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2  & 4  & 8 & 16 & 32 & 18 & 36 & 26  & 6 & 12 & 24  & 2 & --    \\
\hline
3 & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3  & 9 & 27 & 35 & 13 & 39 & 25 & 29 & 41 & 31  & 1  & 3 & 11   \\
\hline
4  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & 16 & 18 & 26 & 12  & 2  & 8 & 32 & 36  & 6 & 24  & 4 & --  \\
\hline
5 & 5 & 25 & 33 & 27 & 43 & 31 & 17 & 39 & 11  & 9 & 45 & 41 & 21 & 13 & 19  & 3 & 15 & 29  & 7 & 35 & 37  & 1  & 5 & 22  \\
\hline
6 & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & 36 & 32  & 8  & 2 & 12 & 26 & 18 & 16  & 4 & 24  & 6 & -- \\
\hline
7 & 7  & 3 & 21  & 9 & 17 & 27  & 5 & 35 & 15 & 13 & 45 & 39 & 43 & 25 & 37 & 29 & 19 & 41 & 11 & 31 & 33  & \textbf{1}  & 7 & 22 \\
\hline
8 & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & 18  & 6  & 2 & 16 & 36 & 12  & 4 & 32 & 26 & 24  & 8 & --  \\
\hline
9 & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 35 & 39 & 29 & 31  & 3 & 27 & 13 & 25 & 41  & 1  & 9 & 11  \\
\hline
10 & 10  & 8 & 34 & 18 & 42  & 6 & 14  & 2 & 20 & 16 & 22 & 36 & 38 & 12 & 28  & 4 & 40 & 32 & 44 & 26 & 30 & 24 & 10 & --  \\
\hline 
11 & 11 & 29 & 43 & 13  & 5  & 9  & 7 & 31 & 19 & 25 & 45 & 35 & 17  & 3 & 33 & 41 & 37 & 39 & 15 & 27 & 21  & 1 & 11 & 22 \\
\hline
12 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12  & 6 & 26 & 36 & 18 & 32 & 16  & 8  & 4  & 2 & 24 & 12 & -- \\
\hline
13 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 31 & 35 & 41 & 27 & 29  & 9 & 25  & 3 & 39  & 1 & 13 & 11  \\
\hline
14 & 14 & 12 & 30  & 6 & 38 & 26 & 42 & 36 & 44 & 18 & 22 & 32 & 34 & 16 & 40  & 8 & 20  & 4 & 10  & 2 & 28 & 24 & 14 & -- \\
\hline
15 & 15 & 41 & 17 & 25  & 7 & 13 & 11 & 27 & 37  & 3 & 45 & 31  & 5 & 29 & 21 & 39 & 33 & 35 & 19  & 9 & 43  & 1 & 15 & 22 \\
\hline
16 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & 26  & 2 & 32  & 6  & 4 & 18 & 12  & 8 & 36 & 24 & 16 & -- \\
\hline
17 & 17 & 13 & 37 & 31 & 21 & 35 & 43 & 41  & 7 & 27 & 45 & 29 & 33  & 9 & 15 & 25 & 11  & 3  & 5 & 39 & 19  & 1 & 17 & 22 \\
\hline
18 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18  & 2 & 36  & 4 & 26  & 8  & 6 & 16 & 12 & 32 & 24 & 18 & -- \\
\hline
19  & 19 & 39  & 5  & 3 & 11 & 25 & 15  & 9 & 33 & 29 & 45 & 27  & 7 & 41 & 43 & 35 & 21 & 31 & 37 & 13 & 17  & 1 & 19 & 22 \\
\hline
20  & 20 & 32 & 42 & 12 & 10 & 16 & 44  & 6 & 28  & 8 & 22 & 26 & 14  & 4 & 34 & 36 & 30  & 2 & 40 & 18 & 38 & 24 & 20 & --  \\
\hline
21 & 21 & 27 & 15 & 39 & 37 & 41 & 33  & 3 & 17 & 35 & 45 & 25 & 19 & 31  & 7  & 9  & 5 & 13 & 43 & 29 & 11  & 1 & 21 & 22  \\
\hline
22 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & 24 & 22 & -- \\
\hline
23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & 23 & -- \\
\hline 
\end{tabular}
\end{center}
}
\caption{Values of $a^i{\rm ~mod~}46, 1\leq a,i<24$}
\label{SrcArith3c}
\label{expmod46}
\end{table}

\vskip +10 pt
\begin{theorem}\label{thm-zth-ordp}
  Given a modulus $n$ and a number $a$, relative prime to $n$ the
  following holds: \\
  The set $\{ a^i\; (\mbox{mod }n) |\; i = 1,\dots,J(n)\}$ equals the
  multiplicative group $Z_n^*$ if and only if $ord_n(a) = J(n)$.\footnote{%
    For prime moduli $p$ all $a$ with $0 < a < p$ are of order $J(p) = p - 1$.
    Compare table~\ref{expmod45} for an example. In this case $a^i (\mbox{mod }n)$ goes
    through all the values $1,\dots,p-1$. Exhausting all possible values of
    the set is an important cryptographic proposition (compare theorem
    \ref{thm-zth-exhperm}). This determines a permutation\index{Permutation}
    $\pi(p-1)$.}$^,$\footnote{%
    Table~\ref{expmod46} demonstrates that for composite moduli $n$ not all $a$ are of
    maximal order $J(n)$. In this example only $5,7,11,15,17,19\mbox{ and }21$
    are of order 22.
  }
\end{theorem}
\noindent The multiplicative group $Z_n^*$ only contains all values from
$1$ to $n-1$, if n is prime (see \ref{def-zth-znmult}).

\newpage
\begin{example}{ 5: Length of cycles}

\noindent The following tables~\ref{expmod14} and~\ref{expmod22}\footnote{%
  See chapter~\ref{nt:AppArith3d}, ``\nameref{nt:AppArith3d}''
  for the source code to generate the tables~~\ref{expmod14} und~\ref{expmod22}
  using Sage.
} serve as samples to introduce cycle lengths\index{cycle length} -- this is a topic
which goes beyond the multiplicative order.

\noindent Cycle here means a sequence of numbers $a^i{\rm ~mod~}n$ with $1\leq i<n$
for a given $a$, and a repeating sequence. Within the cycle each number is unique.
Generally, cycles don't have to contain the $1$ -- in the opposite to cycles
which have a multiplicative order $\ge 1$ (they yet have the $1$ always at the end).\\
With $l$ we low mean the cycle length.

\noindent The maximum cycle length $l_{max}$ is $J(n)$.\\
For the following tables~\ref{expmod14} and~\ref{expmod22} $J(n)$ is
(according to theorem~\ref{J_of_n}):\\
\indent  - $J(14) = J(2*7) = 1*6 = 6$.\\
\indent  - $J(22) = J(2*11) = 1*10 = 10$.

\noindent a) If the multiplicative order exists for $a$, (indendently whether
$a$ is prim) it is: $ord_{n}(a) = l$.
\indent Samples: The maximum length $l_{max}$\footnote{%
            We don't know of a formular telling for which a the length has a maximum.}
is achieved e.g. for:\\
\indent  - $a=3$ with $l_{max} = ord_{14}(a) = 6$ in table~\ref{expmod14}, or\\
\indent  - $a=10$ with $l_{max} = ord_{22}(a) = 10$ in table~\ref{expmod22}.

\noindent b) Also, if no multiplicative order exists for $a$, the maximum cycle
length can be achieved.\footnote{%
         Here the sequences are built via $a^i{\rm ~mod~}n$ with $1\leq i<n$.
         For composite numbers $n$ the sequences never contain all numbers
         $1, ..., n-1$.\\
         This should not be mixed up with RSA, where the ``sequence'' is built
         differently, $m^e{\rm ~mod~}n$ with $0\leq m<n$, and this sequence
         then takes all numbers $0, ..., n-1$ (permutation).
}\\
\indent Samples:\\
\indent  - In table~\ref{expmod14}: $l_{max}=J(14)=6$ for $a=10, 12$.\\
\indent  - In table~\ref{expmod22}: $l_{max}=J(22)=10$ for $a=2, 6, 8, 18$.

\end{example}

\begin{table}[ht]
\begin{center}
\begin{tabular}{|l||c|c|c|c|c|c|c|c|c|c|c|c|c||c|c|c|}
\hline
$a \setminus i$ & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 & $ord_{14}(a)$       & $J(14)$ & $l$ \\
\hline
\hline
1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 6 & 1 \\
\hline
2 & 2 & 4 & 8 & 2 & 4 & 8 & 2 & 4 & 8 & 2 & 4 & 8 & 2 & 0 & 6 & 3 \\
\hline
3 & 3 & 9 & 13 & 11 & 5 & 1 & 3 & 9 & 13 & 11 & 5 & 1 & 3 & 6 & 6 & 6 \\
\hline
4 & 4 & 2 & 8 & 4 & 2 & 8 & 4 & 2 & 8 & 4 & 2 & 8 & 4 & 0 & 6 & 3 \\
\hline
5 & 5 & 11 & 13 & 9 & 3 & 1 & 5 & 11 & 13 & 9 & 3 & 1 & 5 & 6 & 6 & 6 \\
\hline
6 & 6 & 8 & 6 & 8 & 6 & 8 & 6 & 8 & 6 & 8 & 6 & 8 & 6 & 0 & 6 & 2 \\
\hline
7 & 7 & 7 & 7 & 7 & 7 & 7 & 7 & 7 & 7 & 7 & 7 & 7 & 7 & 0 & 6 & 1 \\
\hline
8 & 8 & 8 & 8 & 8 & 8 & 8 & 8 & 8 & 8 & 8 & 8 & 8 & 8 & 0 & 6 & 1 \\
\hline
9 & 9 & 11 & 1 & 9 & 11 & 1 & 9 & 11 & 1 & 9 & 11 & 1 & 9 & 3 & 6 & 3 \\
\hline
10 & 10 & 2 & 6 & 4 & 12 & 8 & 10 & 2 & 6 & 4 & 12 & 8 & 10 & 0 & 6 & 6 \\
\hline
11 & 11 & 9 & 1 & 11 & 9 & 1 & 11 & 9 & 1 & 11 & 9 & 1 & 11 & 3 & 6 & 3 \\
\hline
12 & 12 & 4 & 6 & 2 & 10 & 8 & 12 & 4 & 6 & 2 & 10 & 8 & 12 & 0 & 6 & 6 \\
\hline
13 & 13 & 1 & 13 & 1 & 13 & 1 & 13 & 1 & 13 & 1 & 13 & 1 & 13 & 2 & 6 & 2 \\
\hline
14 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 6 & 1 \\
\hdashline
\hdashline
15 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 1 & 6 & 1 \\
\hline
16 & 2 & 4 & 8 & 2 & 4 & 8 & 2 & 4 & 8 & 2 & 4 & 8 & 2 & 0 & 6 & 3 \\
\hline
\end{tabular}
\end{center}
\caption{Values of $a^i{\rm ~mod~}14, 1\leq a<17, i<14$}
\label{expmod14}
\end{table}
